---
title: "`r params$cohort` — cohort report"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M %Z')`"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    toc_depth: 2
    self_contained: true
    df_print: paged
params:
  cohort: "MoHQ"
  manifest: "manifest.tsv"
  completeness: "completeness.tsv"
  figdir: "figures"
  run_params: "run_params.tsv"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
suppressPackageStartupMessages({
  library(data.table); library(knitr); library(DT); library(htmltools)
})

figs <- if (dir.exists(params$figdir))
  list.files(params$figdir, pattern = "\\.(png|html)$", full.names = TRUE) else character()

# ---------------------------------------------------------------------------
# Figure embedding.
#
# The previous template mixed two mechanisms: htmltools::tags$img(src=...) for
# some panels and knitr::include_graphics() for others. include_graphics() is
# the one that reliably survives self_contained rendering. Standardising on it
# means the whole report travels as a single file rather than half the panels
# breaking once it leaves the working directory.
#
# Interactive HTML (the PCA widget) genuinely cannot be inlined into a parent
# self-contained document via <iframe>; that link WILL break when the file is
# moved. We link it explicitly and say so, instead of embedding a dead frame.
# ---------------------------------------------------------------------------
show_fig <- function(pattern, alt = "figure") {
  hit <- grep(pattern, figs, value = TRUE)
  png <- hit[grepl("\\.png$", hit)]
  if (!length(png)) return(invisible(NULL))
  for (p in png) cat(sprintf("\n![%s](%s)\n\n", alt, p))
  invisible(TRUE)
}

# ---------------------------------------------------------------------------
# Per-figure explanation.
#
# A figure with no caption is read by whoever opens the report months later,
# out of context, usually without the person who made it. `reads` says how to
# interpret the axes; `caveat` says what would make it misleading.
#
# Keep caveats about the METHOD and about how to read the figure -- denominators,
# gene selection, power, what a metric can and cannot show. They are guidance
# for the reader. Do not write pipeline history into them: a delivered report is
# not the place to narrate what was fixed along the way, and a reader who was
# not there reads it as a warning about the current output.
#
# Printed BEFORE the figure, not after: a reader who scrolls past a plot they
# have already misread does not come back for the caption.
# ---------------------------------------------------------------------------
fig_note <- function(title, reads, caveat = NULL) {
  cat(sprintf(
    paste0("\n<div style='margin:14px 0 6px 0;padding:10px 12px;",
           "border-left:4px solid #1C7293;background:#F4F7F9;font-size:0.94em;'>",
           "<strong>%s</strong><br>%s%s</div>\n\n"),
    title, reads,
    if (is.null(caveat)) ""
    else sprintf("<br><span style='color:#8A5A2B;'><strong>Careful:</strong> %s</span>",
                 caveat)))
  invisible(TRUE)
}

# ---------------------------------------------------------------------------
# Missing-panel reporting.
#
# The previous fallbacks said things like "bypassed via user toggles" and
# "skipped during execution optimization workflows" for every absent figure,
# regardless of cause. Reading the report months later you could not tell
# whether a panel was missing because you disabled it, because the process
# crashed, or because the cohort had no such data. Those need different actions,
# so they are reported differently here.
# ---------------------------------------------------------------------------
# The parameter record is staged with the figures, not passed as its own path,
# so look for it in both places. Previously only params$run_params was tried,
# nothing was ever staged there, and the section silently reported that no
# parameters had been recorded -- in the one section whose whole purpose is to
# say what the run used.
.runp_path <- c(params$run_params,
                file.path(params$figdir, "run_params.tsv"))
.runp_path <- .runp_path[file.exists(.runp_path)]
runp <- if (length(.runp_path))
  tryCatch(fread(.runp_path[1], sep = "\t"), error = function(e) NULL) else NULL
param_val <- function(k, default = NA) {
  if (is.null(runp) || !"param" %in% names(runp)) return(default)
  v <- runp[param == k]$value
  if (length(v)) v[1] else default
}

absent <- function(pattern, toggle = NULL, data_reason = NULL) {
  # MATCH IMAGES ONLY, like show_fig does.
  #
  # `figs` now carries the analyses' .tsv tables as well as their .png plots,
  # so a bare grep would treat "the table exists" as "the figure exists" and
  # silently suppress the missing-panel notice for a step that produced its
  # numbers and failed to draw them -- the one case this function is for.
  if (length(grep(pattern, figs[grepl("\\.png$", figs)]))) return(invisible(NULL))
  enabled <- if (!is.null(toggle)) param_val(toggle) else NA
  msg <- if (!is.na(enabled) && tolower(as.character(enabled)) %in% c("false", "0", "no")) {
    sprintf("**Not run.** `%s` was set to `false` for this run.", toggle)
  } else if (!is.null(data_reason)) {
    sprintf("**No output.** %s", data_reason)
  } else {
    paste("**No output, and the step was enabled.** This means the process",
          "failed or produced nothing. Check the Nextflow log for this cohort",
          "before interpreting the rest of this report.")
  }
  cat("\n<div style='padding:10px;border-left:4px solid #d95f02;background:#fff8f0;'>\n",
      msg, "\n</div>\n\n")
}

man  <- tryCatch(fread(params$manifest, sep = "\t", na.strings = c("NA", "")),
                 error = function(e) NULL)
comp <- tryCatch(fread(params$completeness, sep = "\t"), error = function(e) NULL)

read_tbl <- function(pat) {
  f <- list.files(params$figdir, pattern = pat, full.names = TRUE)
  if (!length(f)) return(NULL)
  tryCatch(fread(f[1], sep = "\t"), error = function(e) NULL)
}
dt_show <- function(d, caption = NULL, digits = 3) {
  if (is.null(d) || !nrow(d)) return(invisible(NULL))
  num <- names(d)[vapply(d, is.numeric, logical(1))]
  datatable(d, caption = caption, rownames = FALSE, filter = "top",
            extensions = "Buttons",
            options = list(pageLength = 10, scrollX = TRUE, dom = "Bfrtip",
                           buttons = c("copy", "csv"))) |>
    formatRound(columns = num, digits = digits)
}
```

## What this report covers {.unnumbered}

```{r summary, results='asis'}
if (!is.null(man)) {
  n      <- nrow(man)
  ready  <- sum(man$analysis_ready == "yes", na.rm = TRUE)
  multi  <- sum(man$multi_tumour == "yes", na.rm = TRUE)
  insts  <- paste(sort(unique(man$institution)), collapse = ", ")
  rna    <- sum(!is.na(man$expression_genes))

  # DO NOT claim a single N for the whole report.
  #
  # The first real report said "44 analysis-ready and contribute to every figure
  # below", while the oncoplot showed n = 20 and the copy-number figures n = 36.
  # All three numbers were correct; the sentence was not. Each analysis has its
  # own denominator, because each depends on a different input file being
  # present -- mutation figures need a PCGR VCF, copy-number figures need CNA
  # segments, expression needs RNA.
  #
  # Every figure below states its own n in its title. This block reports the
  # eligible population per analysis so the differences are visible up front
  # rather than discovered by comparing figure titles.
  # EVERY analysis sits behind the analysis_ready gate.
  #
  # main.nf builds ch_ready = rows where analysis_ready == 'yes', and every
  # downstream channel -- ch_cna, ch_cnvkit, the mutation channels, expression,
  # fusions -- branches off it. So holding the input file is necessary but not
  # sufficient: a patient missing ANY required asset never reaches ANY figure.
  #
  # Counting the input column alone overstates all four rows. On MoHQ-MU-16 it
  # printed 86 eligible for mutations while the oncoplot contained 73, because
  # 29 patients lack cna_segments and are gated out of everything, mutations
  # included. Same class of error as the 49-vs-20 one this block already
  # documents -- there the wrong COLUMN, here the missing GATE.
  has <- function(..., ready_only = TRUE) {
    cols <- c(...)
    cols <- cols[cols %in% names(man)]
    if (!length(cols)) return(NA_integer_)
    # A patient counts if ANY of the listed columns is populated.
    ok <- Reduce(`|`, lapply(cols, function(cc) !is.na(man[[cc]]) & man[[cc]] != "NA"))
    if (ready_only && "analysis_ready" %in% names(man))
      ok <- ok & (man$analysis_ready == "yes")
    sum(ok)
  }

  # Gap-fill changes who has a PCGR VCF, and the manifest cannot know it.
  #
  # pcgr_vcf is written before the run. With --run_pcgr_gapfill true, RUN_PCGR
  # creates a PCGR VCF for every `regenerable` patient, so they DO reach the
  # mutation figures -- 16 of them here. Counting only pcgr_vcf therefore
  # understates the mutation row in exactly the situation where the pipeline
  # did the most work.
  gapfill_on <- isTRUE(as.logical(param_val("run_pcgr_gapfill", FALSE)))
  n_mut <- function(col) {
    if (!identical(col, "pcgr_vcf") || !gapfill_on) return(has(col))
    ok <- ((!is.na(man[[col]]) & man[[col]] != "NA") |
           (man$pcgr_vcf_status %in% "regenerable"))
    if ("analysis_ready" %in% names(man)) ok <- ok & (man$analysis_ready == "yes")
    sum(ok)
  }

  # Each row must name the column the analysis ACTUALLY reads, not the one that
  # sounds right. The first version of this table said 49 eligible for mutations
  # and 44 for copy number, while the figures showed 20 and 36 -- because it
  # counted `mutation_vcf` (populated for ensemble fallbacks too, which
  # inhibit_vep excludes) and `cna_segments` (which is the PCGR table, not the
  # CNVkit VCF that BUILD_COHORT_SEG turns into the .seg the CNV figures use).
  #
  # A denominator table that disagrees with the figures is worse than none: it
  # looks authoritative and is wrong.
  mut_col <- if (isTRUE(as.logical(param_val("inhibit_vep", TRUE)))) "pcgr_vcf" else "mutation_vcf"
  mut_req <- if (identical(mut_col, "pcgr_vcf"))
    paste0("PCGR VCF",
           if (gapfill_on) " (delivered, or regenerated by this run)"
           else " (inhibit_vep is on, so ensemble-only patients are excluded)")
  else "any somatic VCF"

  cat(sprintf("**%d patients** in this cohort, **%d analysis-ready**.\n\n", n, ready))
  cat("Each analysis uses only the patients that have its required input, so the denominators differ. Every figure states its own n.\n\n")
  elig <- data.frame(
    Analysis = c("Mutations (oncoplot, TMB)", "Copy number (frequency, burden)",
                 "Expression (PCA)", "Fusions"),
    Requires = c(mut_req, "CNVkit VCF (the cohort .seg is built from these)",
                 "expression file", "Anno-FUSE or LINX calls"),
    `Eligible patients` = c(n_mut(mut_col), has("cnvkit_vcf"),
                            has("expression_genes"), has("anno_fuse", "linx_fusion")),
    check.names = FALSE)
  print(kable(elig))
  cat("\n")
  cat(sprintf("_Counts are patients that hold the input **and** pass the analysis-ready gate (%d of %d), since every analysis reads from that set._\n\n",
              ready, n))
  cat("_If a figure's n differs from the number here, the analysis dropped patients for a further reason -- check its log._\n\n")
  cat(sprintf("- Institution(s): %s\n", insts))
  cat(sprintf("- With RNA: %d/%d\n", rna, n))
  if (multi) cat(sprintf("- %d patient(s) have more than one tumour sample; only the lowest-aliquot tumour is used, so each patient counts once in every frequency.\n", multi))
  if (n > ready) {
    cat(sprintf("\n**%d patient(s) are excluded** from the analyses:\n\n", n - ready))
    ex <- man[analysis_ready != "yes", .(patient_id, missing = missing_core)]
    print(kable(ex))
    cat("\n")
  }
} else {
  cat("_Manifest unavailable._\n")
}
```

### Parameters this run used {.unnumbered}

```{r params_table}
# The previous report showed figures with no record of the thresholds behind
# them. For an archive that will be read years later, the parameters are as
# important as the plots.
if (!is.null(runp)) {
  datatable(runp, rownames = FALSE, options = list(pageLength = 8, dom = "tp"))
} else {
  cat("_No parameter record staged._")
}
```

> **Mutation data provenance.** The MAFs distributed in the MoHQ collection are
> not gene-annotated (`Hugo_Symbol` is `Unknown` throughout). Every mutation
> figure below is built from MAFs regenerated from the PCGR VCFs with
> vcf2maf + VEP inside this pipeline, not from the collection MAFs.

---

## Data completeness

```{r completeness, results='asis'}
fig_note(
  "Completeness — which patients hold which files",
  paste("One row per patient, one column per expected asset, coloured by",
        "present or absent. Reading down a column shows an asset missing across",
        "many patients, which usually points at a processing round rather than",
        "at individual patients; reading across a row shows a patient missing",
        "several things at once."),
  paste("this reflects what was HARVESTED, not necessarily what exists on the",
        "object store. A file set excluded at harvest time appears here as",
        "uniformly absent — check the harvest sets before reporting a gap",
        "upstream."))
show_fig("completeness_heatmap", "Completeness")
absent("completeness", data_reason = "No completeness matrix was produced.")
```

```{r completeness_tbl}
if (!is.null(comp)) {
  ac <- setdiff(names(comp), c("cohort_id", "patient_id", "analysis_ready"))
  s <- data.table(asset = ac,
                  present = vapply(ac, function(a) sum(comp[[a]] == "yes"), numeric(1)),
                  total = nrow(comp))
  s[, pct := round(100 * present / total, 1)]

  # Split the two questions the one table was answering at once. Assets present
  # for NO patient are a harvest-scope choice far more often than a delivery
  # gap, and mixing them into the same list makes the cohort look far emptier
  # than it is. They are also excluded from the heatmap above for the same
  # reason, so the table has to say where they went.
  never <- s[present == 0]
  s     <- s[present > 0]

  # Narrative FIRST, table LAST. Only the last expression in a chunk is
  # auto-printed, so a dt_show() followed by anything else is built and then
  # discarded -- the table vanishes silently.
  if (nrow(never)) {
    cat(sprintf(paste0("\n### Assets absent for every patient (%d)\n\n",
                       "These are **not** in the heatmap above. An asset missing for ",
                       "all %d patients is usually a harvest scope choice rather than ",
                       "a delivery gap — `harvest_juno.py` transfers a chosen `--sets` ",
                       "and everything else is absent locally by design. Confirm ",
                       "against the object store before reporting any of these ",
                       "upstream.\n\n"),
                nrow(never), nrow(comp)))
    print(kable(never[order(asset), .(asset)]))
    cat("\n\n")
  }

  dt_show(s[order(pct)], "Completeness by asset, for assets present in at least one patient",
          digits = 1)
}
```

## Somatic mutations

```{r onco, results='asis'}
fig_note(
  "Oncoplot — the cohort's mutation landscape",
  paste("One row per gene, one column per patient, ordered by how many patients",
        "carry a mutation. Colour is the variant class (missense, nonsense,",
        "frameshift, splice site); black means more than one mutation in that",
        "gene for that patient. The bar above each column is that patient's",
        "mutation burden; the percentage on the right is the fraction of the",
        "cohort altered in that gene. Copy-number gains and losses are overlaid",
        "from PCGR's <code>cna_segments</code>, joined on gene symbol."),
  paste("genes are selected by mutation FREQUENCY, and in whole-genome data that",
        "ranking returns long, repetitive and segmentally duplicated genes",
        "before it returns drivers. An artefact filter is applied (see the",
        "unfiltered ranking table below), but it is a blocklist and blocklists",
        "do not converge — check that the top genes are plausible for this",
        "tumour type before drawing any conclusion."))
show_fig("_oncoprint\\.png", "Oncoplot")
absent("_oncoprint\\.png", toggle = "run_oncoprint")
```

```{r gene_tbl}
dt_show(read_tbl("_gene_summary\\.tsv$"), "Per-gene mutation summary")
```

```{r gene_rank_note, results='asis'}
# TEXT chunk. Markdown from cat() needs results='asis'; the WIDGET below must
# NOT be in an asis chunk, because a DT table needs the normal knit_print path
# and renders as nothing under 'asis'. Hence two chunks rather than one.
#
# The ranking BEFORE the artefact filter, alongside the one after it. Showing
# both lets a reader judge the filter rather than trust it: `is_flags` marks
# every gene removed, so the two lists together say more than either alone.
rk <- read_tbl("_gene_rank_unfiltered\\.tsv$")
if (!is.null(rk) && "is_flags" %in% names(rk)) {
  n_flag <- sum(as.logical(rk$is_flags), na.rm = TRUE)
  if (n_flag > 0) {
    cat(sprintf(paste0("\n**%d of the %d ranked genes** are on the artefact list and ",
                       "do not appear in the oncoplot above. These are long, repetitive ",
                       "or segmentally duplicated genes that accumulate mapping errors ",
                       "and rank high on frequency without being drivers. The `is_flags` ",
                       "column marks them.\n\n"),
                n_flag, nrow(rk)))
  }
}
```

```{r gene_rank_tbl}
dt_show(rk, "Gene ranking before the artefact filter (is_flags = excluded)")
```

```{r oncodrive_note, results='asis'}
od <- read_tbl("_oncodrive\\.tsv$")
if (!is.null(od)) {
  fig_note(
    "Genes ranked by clustering, not by count",
    paste("Frequency ranking favours long genes: a bigger coding sequence is a",
          "bigger target for passenger mutations. This asks a different",
          "question — are a gene's mutations <em>concentrated</em> at",
          "particular residues, more than chance would give? Positive",
          "selection piles mutations on the same few positions; a passenger",
          "process scatters them. Gene length cancels out."),
    paste("this finds hotspot oncogenes (BRAF V600, NRAS Q61, KRAS G12) and",
          "systematically MISSES tumour suppressors — CDKN2A, NF1 and TP53 are",
          "inactivated by truncating mutations spread along the gene, which is",
          "exactly the pattern this test scores as 'not clustered'. Read it",
          "beside the frequency ranking, not instead of it."))
}
```

```{r oncodrive_fig, results='asis'}
show_fig("_oncodrive\\.png", "Genes with clustered mutations")
```

```{r oncodrive_tbl}
dt_show(od, "Positional clustering per gene (oncodrive; low FDR = clustered)")
```

```{r rank_cmp_tbl}
# The two rankings side by side. High on frequency and absent from clustering
# is the length-bias case; high on clustering and modest on frequency is what
# the frequency ranking was hiding.
dt_show(read_tbl("_rank_comparison\\.tsv$"),
        "Frequency rank vs clustering rank, per gene")
```



```{r tmb, results='asis'}
fig_note(
  "Tumour mutational burden — mutations per megabase, per patient",
  paste("Non-synonymous coding variants divided by the callable territory.",
        "Patients are ordered by burden, so hypermutators sit at one end.",
        "TMB is used clinically as an immunotherapy-response marker, which is",
        "why the absolute value matters and not just the ranking."),
  paste("the denominator must match what the numerator counts. This pipeline",
        "counts CODING variants, so the denominator is the coding footprint",
        "(~30 Mb) even though this is whole-genome sequencing. A whole-genome",
        "denominator of 2,800 Mb would scale every value down by roughly 90x,",
        "and the result would still fall in a readable range — so check the",
        "value stated above against the assay rather than the plot."))
show_fig("_tmb\\.png", "Tumour mutational burden")
```

```{r tmb_tbl}
tmb <- read_tbl("_tmb\\.tsv$")
if (!is.null(tmb)) {
  cat(sprintf("\nTMB assumes **%s Mb** of callable territory. ",
              param_val("callable_mb", "?")))
  cat("If that does not match the assay (~30 Mb exome, ~2800 Mb WGS), every TMB value scales proportionally.\n\n")
  dt_show(tmb, "Tumour mutational burden per patient")
}
```

```{r pathways, results='asis'}
fig_note(
  "Oncogenic pathways — mutations grouped by the pathway they act in",
  paste("Genes are collapsed into curated signalling pathways (RTK-RAS, TP53,",
        "cell cycle, and so on) and the fraction of the cohort affected in each",
        "is shown. Useful because two patients can have different genes",
        "mutated and the same pathway disrupted, which a gene-level view hides."),
  paste("pathway membership comes from a fixed curated list, so a gene absent",
        "from that list contributes to nothing — a pathway reading 0% may mean",
        "'not mutated' or 'not annotated'."))
show_fig("_pathways\\.png", "Oncogenic pathways")
# Every other panel has this guard; pathways did not, so a missing figure left
# an explanatory note hanging above nothing.
absent("_pathways\\.png", toggle = "run_oncoprint",
       data_reason = paste("The pathway table below was produced but the plot was not.",
                           "maftools draws it as a side effect, so this usually means",
                           "no gene in the cohort mapped to a curated pathway."))
```

```{r pathway_tbl}
dt_show(read_tbl("_pathways\\.tsv$"), "Oncogenic pathways: fraction of the cohort affected")
```

```{r interactions, results='asis'}
fig_note(
  "Mutual exclusivity and co-occurrence",
  paste("Every pair of frequently mutated genes tested for whether they occur",
        "together more or less often than chance. Mutual exclusivity can point",
        "to two routes to the same phenotype (only one is needed);",
        "co-occurrence can point to cooperation, or to a shared cause such as",
        "a hypermutator phenotype."),
  paste("this tests many gene pairs at once on a small cohort, so with n in the",
        "tens most cells hold single-digit counts. Treat significance here as a",
        "hypothesis, not a result."))
show_fig("_somatic_interactions", "Mutual exclusivity / co-occurrence")
```


```{r interactions_tbl}
dt_show(read_tbl("_somatic_interactions\\.tsv$"),
        "Gene-pair mutual exclusivity / co-occurrence tests")
```

## Copy number

```{r cnvfreq, results='asis'}
fig_note(
  "Genome-wide copy-number frequency",
  paste("The genome is divided into 1 Mb bins and, for each bin, the percentage",
        "of patients with a gain is drawn above the axis and the percentage",
        "with a loss below it, left to right across the chromosomes. Peaks are",
        "regions recurrently altered across the cohort — the starting point for",
        "asking which gene in that region is being selected for."),
  paste("this shows WHERE alterations recur, not whether they matter. Broad arm",
        "and chromosome-level events dominate the picture, and identifying a",
        "driver within a peak needs a focal-peak method such as GISTIC, which",
        "is a separate step."))
show_fig("_cnv_frequency", "Genome-wide CNV frequency")
absent("_cnv_frequency", toggle = "run_cnv_frequency")
```

```{r burden, results='asis'}
fig_note(
  "Copy-number burden — Fraction of Genome Altered",
  paste("Per patient, the proportion of the profiled genome that is gained or",
        "lost. A single-number summary of genome instability: near zero means a",
        "quiet genome, high values mean widespread rearrangement. Patients",
        "profiled over too little of the genome are marked, and values at or",
        "above 99% are marked separately."),
  paste("the denominator is the length actually profiled FOR THAT PATIENT, not",
        "a fixed genome size, so values stay comparable between samples with",
        "different callable footprints. Check <code>profiled_mb</code> in the",
        "table below alongside any value you rely on. An FGA at or above 99%",
        "indicates a copy-number normalisation problem for that sample rather",
        "than an exceptionally unstable tumour."))
show_fig("_cnv_burden\\.png", "Copy-number burden")
absent("_cnv_burden\\.png", toggle = "run_cnv_burden")
show_fig("_by_institution", "FGA by institution")
```

```{r fga_tbl}
# ORDER MATTERS: only the LAST expression in a chunk is auto-printed, and a DT
# widget is an object like any other. A dt_show() followed by a cat() creates
# the widget and then throws it away -- the table simply does not appear, with
# no error. So any narrative text goes FIRST and dt_show() goes LAST.
fga <- read_tbl("_cnv_burden\\.tsv$")
if (!is.null(fga)) {
  if ("low_coverage" %in% names(fga) && any(fga$low_coverage)) {
    cat(sprintf("\n**%d patient(s) flagged low-coverage** (profiled over less than %s Mb of autosome). Their FGA is not comparable to the rest of the cohort.\n\n",
                sum(fga$low_coverage), param_val("min_profiled_mb", "?")))
  }
  dt_show(fga, "Fraction of Genome Altered per patient")
}
```

> FGA is altered autosomal length divided by the length **actually profiled for
> that patient**, not by a fixed genome size. `profiled_mb` in the table above
> shows that footprint.

## Expression

```{r pca, results='asis'}
fig_note(
  "Expression PCA — do samples separate, and by what?",
  paste("Each point is a patient, positioned by the two directions of largest",
        "variation across the most variable genes. Points close together have",
        "similar expression profiles. The scree plot beneath shows how much",
        "variance each component captures — if PC1 holds most of it, one",
        "dominant factor is driving the data."),
  paste("this figure is diagnostic before it is biological. The question to ask",
        "first is whether separation tracks a TECHNICAL variable — processing",
        "round, institution, PCGR version — because if it does, downstream",
        "comparisons are measuring the pipeline. Colouring by a constant (such",
        "as institution within a single-institution cohort) gives one colour",
        "and answers nothing."))
show_fig("_expression_pca\\.png", "Expression PCA")
absent("_expression_pca\\.png", toggle = "run_pca",
       data_reason = "This cohort has fewer than three RNA samples, so a PCA cannot be computed.")
show_fig("_pca_scree", "Scree plot")

# The interactive widget cannot be inlined into a self-contained parent
# document. Link it and be explicit, rather than embedding an iframe that
# silently breaks the moment the report is copied elsewhere.
ih <- grep("_expression_pca\\.html$", figs, value = TRUE)
if (length(ih)) {
  cat(sprintf("\n[Open the interactive PCA](%s) — separate file, must travel alongside this report.\n\n", ih[1]))
}
```

```{r pca_out}
sc <- read_tbl("_pca_scores\\.tsv$")
if (!is.null(sc) && "is_outlier" %in% names(sc)) {
  o <- sc[is_outlier == TRUE]
  if (nrow(o)) {
    cat(sprintf("\n**%d expression outlier(s)** (robust distance > 4 MAD on PC1/PC2):\n\n", nrow(o)))
    print(kable(o[, .(patient_id, PC1, PC2, outlier_score)], digits = 2))
  }
}
```

> Points are coloured by a real covariate, not by a constant. Separation that
> tracks institution rather than biology is a batch effect and should be
> resolved before the downstream results are interpreted.

## Fusions

```{r fusions, results='asis'}
fig_note(
  "Recurrent fusions — gene pairs seen in more than one patient",
  paste("Fusion calls aggregated across the cohort and ranked by how many",
        "patients carry each gene pair. Two independent sources are used:",
        "Anno-FUSE from the RNA, and LINX from the whole-genome structural",
        "variants. A pair supported by BOTH has RNA evidence of a transcript",
        "and DNA evidence of the rearrangement that would produce it, which is",
        "much stronger than either alone."),
  paste("recurrence is the signal, and a fusion in a single patient is a",
        "curiosity rather than a finding. Read-through transcripts between",
        "neighbouring genes are a common false positive in RNA-only calls —",
        "if a pair has no DNA support, check whether the two genes are",
        "adjacent."))
show_fig("_fusions", "Recurrent fusions")
absent("_fusions", toggle = "run_fusions",
       data_reason = "No fusion calls were available for this cohort.")
```

```{r fusion_tbl}
dt_show(read_tbl("_fusions\\.tsv$"), "Recurrent fusions (DNA+RNA supported flagged)")
```

## Stratified comparison

```{r comp, results='asis'}
fig_note(
  "Stratified comparison — two groups, same cohort",
  paste("The cohort split by a clinical variable and compared four ways: a",
        "side-by-side oncoplot, a per-gene test of mutation frequency between",
        "groups, gene-level copy-number differences, and genome instability",
        "(FGA) by group. Sex chromosomes are excluded, because when stratifying",
        "by sex every gene on X or Y separates perfectly for reasons of",
        "karyotype rather than tumour biology."),
  paste("genes are chosen from this same cohort's mutation frequencies and then",
        "tested on it, so the p-values are optimistic. Tests are labelled",
        "unadjusted unless a covariate actually varies — within a single",
        "institution, nothing is being adjusted for. Read a null result here as",
        "'underpowered' rather than 'no difference'."))
show_fig("_comparative", "Stratified comparison")
absent("_comparative", toggle = "run_comparative",
       data_reason = "Not enough patients per group for a meaningful comparison.")
```

```{r comp_caveat, results='asis'}
if (length(grep("_comparative", figs))) {
  st <- read_tbl("_cnv_stats\\.tsv$")
  if (!is.null(st) && "gene_selection" %in% names(st)) {
    cat(sprintf("\n> **Gene selection:** %s\n\n", st$gene_selection[1]))
  }
  cat("> Where genes were chosen from this same cohort's mutation frequencies and then tested on it, the p-values are optimistic. Unadjusted tests are labelled as such; institution-adjusted models are reported alongside where the data support them.\n\n")
  dt_show(st, "Per-gene CNV frequency differences")
}
```

## Provenance

```{r prov, results='asis'}
v <- list.files(params$figdir, pattern = "versions", full.names = TRUE)
if (length(v)) {
  cat("```\n")
  for (f in v) cat(paste(readLines(f, warn = FALSE), collapse = "\n"), "\n")
  cat("```\n")
} else cat("_No version files staged._\n")
```

```{r si}
sessionInfo()
```
